Canopy leaf area of an evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability

### 2015-05-19 manuscript.Rmd
Teresa's edits

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [CO~2~] but tracks water availability
author: ''
date: ''
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].  Ecosystem models generally predict that *L* increases with eC~a~, largely in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012]. Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].

@@

## Site description and experimental design
The EucFACE experiment is located on an ancient alluvial floodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l. elevation (33^o^ 37'S, 150^o^ 44'E). It is located within a remnant patch  of native Cumberland Plain woodland [@Tozer2003]. The vegetation, maintained in this location since the 1880’s, retains some old-growth trees and some re-growth vegetation though it has been undisturbed for ca. 70 years. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au). The potential evapotranspiration (PET) is ca. 1500mm (*to be refined!*), giving a ratio of P/PET of ca. 0.53 (*quite dry - details to be added*).

\
The soil at the site is a loamy sand in the Clarendon Formation [@Barton2010; @Isbell2002], a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees ha^-1^. The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

@@

## *L* from diffuse transmittance
In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height similar to the approach used by Norby et al. [-@Norby2003]. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct. 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged in 30 min time steps. At a nearby site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy. For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98. We further trimmed the data by excluding low solar elevation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

@@

where $\tau_p$ canopy transmission estimated from canopy photos. We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests. Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)`% higher than $L_p$ (Fig. SI2, R^2^=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig. S2). We also estimated the extinction coefficient that minimized the difference between the two estimates of *L* (Fig. S2), giving *k* = `r kopt`. These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

@@

\
We found no differences in the *L* anomaly between the C~a~ treatments, as shown by consistent overlap of the the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any timepoint, Fig. 3). A brief exception was the period where one of the rings was affected by an insect outbreak (see above). 
A linear mixed-effects model with C~a~ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on the *L* anomaly (p `r pval(LanomAOV[[3]][1])`). What difference between C~a~ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was `r LanomCIamb` m^2^ m^-2^ for aC~a~, and `r LanomCIele` m^2^m^-2^ for the eC~a~ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` m^2^m^-2^).

@@

##Continuous measurement of *L*
Studying canopy leaf area index in an evergreen natural woodland poses unique challenges. Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan (LL) is unknown (instead, we used *L* and litter production to estimate LL). We can also not rely on leaf area estimates from allometric relationships with stem diameter or height, since these equations are not available for mature *E. tereticornis*, and would be useless anyway since *L* is highly variable temporally (Figs. 1, 7). We found that indirect estimates of *L* based on canopy transmittance were very reliable because relatively small temporal changes in *L* could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs. 1, 5, 7). We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999]. As expected, when we used unfiltered canopy transmittance data, the seasonal patterns in canopy transmittance were much more distinct (Fig. S3), because longer path lengths in the winter result in lower overall transmittance. Using diffuse PAR only avoids having to correct for these path length effects.

>>>>

---
title: Canopy leaf area of an evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability
author: ''
date: ''
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].  Ecosystem models generally predict that *L* increases with eC~a~, largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012]. Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].

@@

## Site description and experimental design
The EucFACE experiment is located on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.). The site is part of a remnant patch  of native Cumberland Plain woodland [@Tozer2003]. The vegetation consists of a native woodland undisturbed for ca. 70 years and maintained in this location since the 1880’s. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au). The potential evapotranspiration (PET) is ca. 1500mm (*to be refined!*), giving a ratio of P/PET of ca. 0.53 (*quite dry - details to be added*).

\
At the site, the soil at the surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots). There is one single dominant species (*Eucalyptus tereticornis* Sm.) with some scattered individuals of *E. amplifolia* Naudin. The understory stratum is  dominated by the grass *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

@@

## *L* from diffuse transmittance
In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct. 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged in 30 min time steps. At a nearby site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy. For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98. We further trimmed the data by excluding low solar e levation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

@@

where $\tau_p$ canopy transmission estimated from canopy photos. We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests. Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)`% higher than $L_p$ (Fig. SI2, R^2^=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig. S2). We also estimated the extinction coefficient that minimized the difference between the two estimates of *L* (Fig. S2), giving *k* = `r kopt`. These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq. 2), weighted towards the zenith.

@@

\
We found no differences in the *L* anomaly between the C~a~ treatments, as shown by consistent overlap of the the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any timepoint, Fig. 3). A brief exception was the period where one of the rings was affected by an insect outbreak (see above). A linear mixed-effects model with C~a~ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on the *L* anomaly (p `r pval(LanomAOV[[3]][1])`). What difference between C~a~ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was `r LanomCIamb` m^2^m^-2^ for aC~a~, and `r LanomCIele` m^2^m^-2^ for the eC~a~ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` m^2^m^-2^).

@@

## Continuous measurement of *L*
Studying canopy leaf area index in an evergreen natural woodland poses unique challenges. Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan (LL) is unknown (instead, we used *L* and litter production to estimate LL). We can also not rely on leaf area estimates from allometric relationships with stem diameter or height, since these equations are not available for mature *E. tereticornis*, and would be useless anyway since *L* is highly variable temporally (Figs. 1, 7). We found that indirect estimates of *L* based on canopy transmittance were very reliable because relatively small temporal changes in *L* could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs. 1, 5, 7). We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999]. As expected, when we used unfiltered canopy transmittance data, the seasonal patterns in canopy transmittance were much more distinct (Fig. S3), because longer path lengths in the winter result in lower overall transmittance. Using diffuse PAR only avoids having to correct for these path length effects.
